\section{GitLab Pipelines}

\begin{lstlisting}[language=yaml]
default: { image: golang:1.24-alpine }    # fallback image for all jobs
stages: [build, test, release, deploy, post-deploy]

build:                       # DevOps cycle -> Build
  stage: build
  script:
    - go mod download
    - CGO_ENABLED=0 go build -o bin/app ./...
  artifacts: { paths: [bin/app] }         # persist for later stages

lint-test:                   # DevOps cycle -> Test
  stage: test
  image: golangci/golangci-lint:v1.64.5-alpine
  script: [golangci-lint run -v]

build-docker:                # DevOps cycle -> Release (publish image)
  stage: release
  image: docker:28
  services: [{ name: docker:28-dind, alias: docker }]   # dind
  variables: { IMG: $CI_REGISTRY_IMAGE:$CI_COMMIT_SHORT_SHA }
  before_script:
    - docker login -u $CI_REGISTRY_USER -p $CI_REGISTRY_PASSWORD $CI_REGISTRY
  script:
    - docker build -t $IMG -f Dockerfile.prod .
    - docker push $IMG

deploy-prod:                 # DevOps cycle -> Deploy
  stage: deploy
  image: alpine/kubectl:1.35
  script:
    - yq -i ".spec.template.spec.containers[0].image=\"$IMG\"" k8s/deployment.yaml
    - kubectl apply -f k8s/deployment.yaml
  environment: { name: production, url: https://demo.example.com }
  rules: [{ if: '$CI_COMMIT_BRANCH == "main"' }]   # prod only on main

post-deploy:                 # DevOps cycle -> Operate/Monitor
  stage: post-deploy
  script: [echo "smoke test + health check"]
\end{lstlisting}

\subsection{Core Concepts \& DSL Mapping}
Pipeline = automated steps $\rightarrow$ faster, safer, reliable. Traceable logs, parallel jobs, consistent envs kill "works on my machine".

\textbf{5-stage flow:} commit/MR/tag $\rightarrow$ \textit{build} (compile, image, package) $\rightarrow$ \textit{test} (unit/integration/security/perf) $\rightarrow$ \textit{release} (publish to registry) $\rightarrow$ \textit{deploy} (target env) $\rightarrow$ \textit{post-deploy} (smoke, health, rollback). \\
\textbf{Map to DevOps cycle:} the pipeline automates the loop's middle -- \texttt{build}$\to$Build, \texttt{test}$\to$Test, \texttt{release}$\to$Release, \texttt{deploy}$\to$Deploy; \textit{triggered by} Code (commit/MR/tag), \textit{feeds} Operate/Monitor. \textbf{CI} = build+test; \textbf{CD} = release+deploy.

\subsection{Jobs, Stages \& Rules}
\texttt{.gitlab-ci.yml} YAML DSL. \textbf{Job} = smallest unit, belongs to one \textbf{stage}, runs \texttt{script} on a runner. Same stage $\rightarrow$ \textit{parallel}; stages $\rightarrow$ \textit{sequential}. \\
\textbf{rules:} \textit{when} a job runs (modern replacement for \texttt{only/except}). \textbf{needs:} builds DAG, start before stage finishes. \textbf{dependencies:} which jobs' artifacts to pull (no ordering). \textbf{workflow:rules:} gate the whole pipeline.
\begin{lstlisting}[language=yaml]
workflow:
  rules:
    - if: $CI_PIPELINE_SOURCE == "merge_request_event"
    - if: $CI_COMMIT_BRANCH && $CI_OPEN_MERGE_REQUESTS
      when: never
    - if: $CI_COMMIT_BRANCH
deploy:
  stage: deploy
  needs: [build]           # DAG, skip stage wait
  dependencies: [build]    # pull build's artifacts
  rules:
    - if: '$CI_COMMIT_BRANCH == "main"'
    - when: never
\end{lstlisting}

\subsection{Cache, Artifacts \& Services}
\textbf{Cache:} shared across jobs/pipelines, speed up repeat work (deps, compiled objects). \\
\textbf{Artifacts:} files persisted to GitLab for downstream stages or download; declare via \texttt{dependencies}/\texttt{needs}. \\
\textbf{Services:} sidecar containers (DB, dind, API) reachable via \texttt{alias}. \\
\begin{lstlisting}[language=yaml]
integration-test:
  services: [{name: postgres:17, alias: db}]
  variables: {POSTGRES_DB: app, POSTGRES_USER: u, POSTGRES_PASSWORD: p}
  script: [psql -h db -U $POSTGRES_USER -d $POSTGRES_DB -c "SELECT 1"]
  cache: {key: deps, paths: [.cache/]}
  artifacts: {paths: [bin/], expire_in: 1h}
\end{lstlisting}

\subsection{Environments}
Where code is deployed. \textbf{Tiers:} Local (workstation) $\rightarrow$ Dev (sandbox) $\rightarrow$ Integration (CI target) $\rightarrow$ Test/QC (interface tests) $\rightarrow$ Staging (prod mirror) $\rightarrow$ Production (live). Linkable to K8s cluster (set up via UI). \texttt{environment.on\_stop} unstages; \texttt{action: stop} stops it.

\subsection{Push vs. Pull Deployment}
\textbf{Push} (Jenkins/GitLab calls cluster): \textcolor{ForestGreen}{\textbf{pro:}} easy, flexible targets (VMs/containers, not just K8s); \textcolor{red}{\textbf{con:}} firewall open, admin access out, pipeline per env.  \\
\textbf{Pull} (agent in cluster, e.g. Argo/Flux): \textcolor{ForestGreen}{\textbf{pro:}} secure (no inbound), auto-reconcile, scales identically; \textcolor{red}{\textbf{con:}} agent per cluster, polling latency.

\subsection{GitLab Runner}
Worker executing jobs defined in .gitlab-ci.yml. \textbf{Scope:} Instance (shared, GitLab-managed) / Group / Project (dedicated). \textbf{Executors:} \texttt{shell}, \texttt{ssh}, \texttt{docker}, \texttt{docker+machine}, \texttt{kubernetes}, \texttt{virtualbox}, \texttt{parallels}. Register: \texttt{gitlab-runner register} with URL + token + tags + executor. Jobs pick runner by \texttt{tags:}.

\subsection{More Keywords \& Predefined Vars}
\texttt{default:} fallback \texttt{image}/\texttt{before\_script}. \texttt{before\_script}/\texttt{after\_script} wrap every \texttt{script}. \texttt{rules:changes:} run only if files changed. \\
\textbf{Predefined vars:} \\
\texttt{\$CI\_REGISTRY[\_IMAGE]}, \\
\texttt{\$CI\_REGISTRY\_USER}/\texttt{\_PASSWORD}, \\
\texttt{\$CI\_COMMIT\_BRANCH}, \\
\texttt{\$CI\_COMMIT\_SHORT\_SHA}, \\
\texttt{\$CI\_PROJECT\_DIR}, \\
\texttt{\$CI\_JOB\_TOKEN} (auth API/registry),
\texttt{\$CI\_MERGE\_REQUEST\_IID}. Pin images by digest (\texttt{img@sha256:..}), not \texttt{:latest} $\rightarrow$ flaky.
