\section{GitOps}
GitOps uses Git repositories as the single source of truth for infrastructure and application configurations. Automated agents continuously monitor Git repos and sync changes to target environments automatically. \\
\textbf{Principles:} \textit{declarative} (desired state lives in Git), \textit{versioned \& immutable} (Git history = audit log, revert = rollback), \textit{pulled automatically} and \textit{continuously reconciled} (agent detects and corrects drift).

\subsection{Push- vs. Pull-Based}
\textbf{Push (classic CI/CD):} the pipeline holds cluster credentials and pushes manifests into the cluster (\texttt{kubectl apply}). - credentials live outside the cluster, - no drift detection, - every new environment needs pipeline changes. \\
\textbf{Pull (GitOps):} an agent inside the cluster watches Git and pulls the desired state. + no external credentials / open firewall, + automatic drift correction, + scales to many clusters, - an agent must be installed per cluster.

\subsection{Drift \& Reconciliation}
\textit{Drift} = the live cluster state diverges from Git (e.g. a manual \texttt{kubectl edit}). The agent runs a \textit{reconciliation loop}: continuously compare Git vs. live state and either report the difference or self-heal (sync/prune) back to the declared state.

\subsection{Argo CD vs. Flux CD}
\textbf{Argo CD:} platform-engineering tool with a web UI \& dashboard. Wraps a deployment in an \texttt{Application} CRD (points at repo+path and a target cluster/namespace). \texttt{ApplicationSet} templates many Applications (multi-cluster, monorepo, per-env). Supports sync waves, health checks, manual or auto sync. \\
\textbf{Flux CD:} lightweight, CLI-first, built from the GitOps Toolkit controllers. Kustomize- \& Helm-native (\texttt{Kustomization}, \texttt{HelmRelease} CRDs), no built-in UI. Strong on image automation \& notifications.

\subsection{Argo CD in Practice}
\begin{lstlisting}[language=sh]
kubectl create namespace argocd
kubectl apply -n argocd -f .../install.yaml
kubectl port-forward -n argocd svc/argocd-server 8080:443
# initial admin password (user: admin):
kubectl -n argocd get secret argocd-initial-admin-secret \
  -o jsonpath="{.data.password}" | base64 -d
\end{lstlisting}
Destination for the local cluster: \texttt{https://kubernetes.default.svc}. \\
\textit{Gotcha:} if a manifest sets \texttt{replicas} \textit{and} an HPA is active, Argo CD and the HPA fight (revert vs rescale) $\rightarrow$ omit \texttt{replicas} from Git. \\
\textbf{Secrets in GitOps:} (1) \textit{encrypted in Git} (Sealed Secrets, SOPS), or (2) \textit{reference in Git} + fetch from an external store (External Secrets, Vault). \\
\textit{Access direction:} push = runner needs access \textit{into} the cluster; pull = the cluster reaches \textit{out} to Git (nothing reaches in).
