\section{Helm}
Package manager for K8 (apt/brew-like).
\textbf{Chart:} files describing related K8 resources.
\textbf{Release:} running instance.
\textbf{Repo:} stores versioned charts.
\textbf{Architecture:} client-side only (Helm 3, no Tiller); talks to K8 API directly; release state stored in-cluster as \textbf{Secrets} (in the release's namespace).
\textcolor{ForestGreen}{\textbf{Pros:}} huge chart library, versioning/deps, lifecycle (install/upgrade/rollback), powerful Go-template logic.
\textcolor{red}{\textbf{Cons:}} templates hard to debug, ``spaghetti YAML'', opaque state, steep curve. Use for: distributing complex apps, consuming community charts, advanced logic. Avoid for: simple YAML tweaks.

\subsection{Commands}
% Placeholders are the exact arg names from helm.sh docs, one color per token.
\begin{lstlisting}[language=sh, escapeinside={@}{@}]
helm repo add|list|rm @\pName{NAME}@ URL    # repos
helm search repo @\pName{<r>}@ / @\pChart{<c>}@ -l                 # all versions
helm install -f vals.yaml @\pName{NAME}@ @\pChart{CHART}@ --version 1.2.3
helm upgrade @\pRelease{RELEASE}@ @\pChart{CHART}@
helm rollback @\pRelease{RELEASE}@ REVISION
helm uninstall @\pRelName{RELEASE\_NAME}@   # also drops history
helm list; helm history @\pRelName{RELEASE\_NAME}@
helm status @\pRelName{RELEASE\_NAME}@      # unknown|deployed|uninstalled|
  # superseded|failed|uninstalling|pending-{install,upgrade,rollback}
helm get all|values|notes|manifest @\pRelName{RELEASE\_NAME}@  # values=overrides
helm template @\pName{NAME}@ @\pChart{CHART}@ -f v.yaml --set k=v  # render local
helm install @\pName{NAME}@ @\pChart{CHART}@ --dry-run --debug     # simulate
helm lint CHART_PATH [--with-subcharts]
helm show chart|readme|values|all @\pChart{CHART}@
helm package CHART_PATH        # build .tgz, then helm push <pkg> <repo>
helm dependency update @\pChart{CHART}@    # pulls into charts/
\end{lstlisting}

\subsection{Chart Structure \& Chart.yaml}
\texttt{helm create <n>} scaffolds: \texttt{Chart.yaml} (meta), \texttt{values.yaml} (defaults), \texttt{templates/} (Go-tmpl manifests), \texttt{templates/\_helpers.tpl} (named templates, \texttt{\_}=not rendered), \texttt{templates/NOTES.txt} (rendered post-install), \texttt{charts/} (subcharts), \texttt{crds/}, \texttt{.helmignore} (globs, \texttt{!}=negate; excludes from \texttt{helm package}).\\
\texttt{Chart.yaml} fields: \textbf{apiVersion} (\texttt{v2}=Helm 3+, \texttt{v1}=legacy), \textbf{name}, \textbf{version} (SemVer2; pkg name = name+ver), \textbf{type} (\texttt{application} default \textbar{} \texttt{library} = helpers only, no install), \textbf{appVersion} (app being shipped, info-only), \textbf{kubeVersion} (range like \texttt{>=1.12.1 <1.14.0 || >=1.14.2}; install fails if mismatch), \textbf{dependencies} (name/version/repository/condition/alias; \texttt{requirements.yaml} deprecated, lockfile on update).

\subsection{Templating}
\textbf{Top-level objects:} \texttt{.Release.\{Name,Namespace,Revision,IsInstall,IsUpgrade,Service\}}, \texttt{.Chart.*}, \texttt{.Values}, \texttt{.Files.Get <path>}, \texttt{.Capabilities.\{APIVersions,KubeVersion\}}, \texttt{.Template.\{Name,BasePath\}}. \\
\textbf{Values precedence} (low$\to$high, last wins): subchart \texttt{values.yaml} $\to$ parent \texttt{values.yaml} $\to$ \texttt{-f myvals.yml} $\to$ \texttt{--set k=v}. Unset default: \texttt{--set x=null}. \\
\textbf{Functions:} Sprig lib (\textasciitilde 60+: \texttt{upper quote eq default trunc toYaml indent nindent}). Pipelines \texttt{|} pass result as last arg. \texttt{include} = \texttt{template} but pipeable. \texttt{required "msg" .Values.x} fails if empty. \texttt{tpl <str> .} evaluates string as template. \\
\textbf{Flow:} \texttt{\{\{ if/else if/else \}\}...\{\{ end \}\}} (false on 0/""/nil/empty); \texttt{\{\{ range \$k,\$v := .Values.list \}\}}; \texttt{\{\{ with .Values.sub \}\}} rescopes \texttt{.}. Vars: \texttt{\{\{- \$n := .Release.Name -\}\}}. \\
\textbf{Whitespace:} \texttt{\{\{-} strips left, \texttt{-\}\}} strips right; \texttt{nindent N} = newline + indent. YAML indent critical.
\begin{lstlisting}[language=yaml]
{{- define "app.name" -}}{{ default .Chart.Name .Values.name | trunc 63 }}{{- end }}
name: {{ include "app.name" . | quote }}
{{- required "set .Values.img!" .Values.img }}
resources: {{- toYaml .Values.resources | nindent 4 }}
\end{lstlisting}
